Attributions for Poverty in Post-Socialist Countries

### 2014-03-08 article2013.Rmd
Lähteet ja pieniä viilauksia
Tsekkasin tekstin sisäiset viitemerkinnät ja lähdeluettelon lehden asetusten mukaisesti. Lähdeluettelossa sulle vielä muutama lisättävä, ne on merkattu lähdeluetteloon.

As Joseph Schumpeter (1942: 12) once argued, "attitudes are coins that do not readily melt". This means that normative expectations, justice beliefs or welfare perceptions are very hard to change. In this way, welfare attitudes can be associated with policy changes by highlighting public sentiments, which are normative background assumptions or ideas that constrain decision-making and institutional change by limiting the range of programme that decision-making elites are likely to perceive as acceptable and legitimate to both their constituents and themselves (Campbell, 2004: 96-100). On the other hand, "the worlds on welfare" literature on welfare attitudes has in turn emphasized that type of welfare policy affects public attitudes (Andress and Heien, 2001; Arts and Gelissen, 2001; Jaeger, 2006; Svallfors, 1997). In regard to public perceptions of the causes of poverty, in the marginal welfare-policy dominated liberal regime, the poor will be asked to fulfil much harder deservingness criteria than in a universal social democratic regime (van Oorschot 2006). In countries, in which welfare programmes are more marginal and less generous, explanations of poverty are more often based on individual morality, reflecting the strength of the dominant individualistic ideology, than they are in the Continental and Nordic welfare states – where external and structural explanations prevail (Albrekt Larsen, 2006: 47-56; Alesina and Gleaser, 2004).

The focus of this study is on the public perceptions of the causes of poverty in post-socialist countries. Since the early 1990s, the development of post-socialist welfare state has received abundant scholarly attention (e.g. Cerami and Vanhuysse, 2009, Inglot, 2008, Haggard and Kauffman, 2008; Cook, 2007; Aidukaite, 2004; Deacon 1992). As the decades under socialist regime have had an undeniable influence on societies, these studies have emphasized the role of institutional differences within region as well as pre-socialist historical origins of the welfare states. Now after 20 years of transition, as there has not been a visible convergence towards a common "post-socialist model", the attention has shifted even more on various institutional and historical explanations (Szikra and Tomka, 2009, Tomka, 2006, Cerami and Vanhuysse, 2009, Inglot, 2008).

The purpose of this article is to examine cross-national differences in attributions for poverty in Europe and Western Asia with special emphasis on post-socialist countries. Prior findings of the comparative studies on public perceptions of the causes of poverty have emphasized clear differences between Western and Eastern European countries (Gallie and Paugam, 2002; Kallio and Niemelä, 2013; Lepianka, 2007; van Oorschot and Halman, 2000). East Europeans are more inclined to blame either the individual behavior or the flaws of the economic system of country than their Western counterparts. Hence, it seems that unlike in Western Europe traditionally opposite attitudinal dimensions, i.e. structural and individualistic explanations of poverty coincide with each other in East Europeans' perceptions. However, there is shortage of comparative studies on attributions for poverty in particular focusing on the East Central European countries or on the whole post-socialist space. There is therefore a need for a more detailed and up-to-date analysis of the public perceptions of the causes of poverty in post-socialist countries.

The novelty of this study is that it covers wide range of countries from Europe and Western Asia. Thus, compared to previous studies, it provides more valid evidence on the differences in perceptions between the East and the West and within post-socialist regime as well as on the importance of the post-socialist legacy. In addition, the study applies multilevel techniques in order to analyze the role of individual and contextual level explanatory factors as determinants of poverty attributions. The article explores to what extent specific risks related to household financial circumstances at the individual level and the state of democracy, income inequality and changes in national economy at the contextual level are associated with different types of lay explanations of poverty. Finally, extensive country-selection of our data provides a novel opportunity to examine the effects of the financial crisis both at the individual and contextual level and their association with attributions for poverty. At the individual level, the study focuses not only on the individuals’ current financial situation. Instead, the article provides an empirical example of how past experiences, current financial position as well as future expectations concerning individual’s economic situation are associated with perceptions of the causes of poverty.

@@

Pioneering empirical studies on lay explanations of poverty emerged at the early 1970s when Joe R. Feagin (1972; 1975) examined Americans' perceptions of the causes of poverty. He categorized reasons for poverty into three basic categories: individualistic, structural and fatalistic. Later, many factor analytic studies have given empirical support to Feagin’s categorization (e.g. Feather, 1974; Furnham, 1982; Hunt, 1996; Morçöl, 1997; Nasser et al., 2002; Niemelä, 2008). Yet some studies have expanded our understanding by incorporating more contemporary beliefs into the attributional scales. These works emphasize especially the relevance of cultural attributions for poverty (Bullock et al., 2003; Cozzarelli et al., 2001).

@@

Prior empirical findings on attributions for poverty in Europe have shown that there are some differences between the West and the East of Europe. Western Europeans are more inclined to support social explanations for poverty than their Eastern counterparts. On the other hand, individual blame type of explanation is generally more popular in the East than in the West of Europe (Kallio and Niemelä, 2013; Kreidl, 2000; Lepianka, 2007; Lepianka et al., 2010; van Oorschot and Halman, 2000). Earlier results suggest, however, that there are variations both within and across groups of countries. Interestingly enough, the variation is highest within the East Central European countries (Kallio and Niemelä, 2013). Also Lepianka et al. (2010) come to the conclusion that further investigation is required due to rather large variation across countries. They showed that individual fate type of explanation is supported relatively less frequently in the East of Europe than in Western European countries. Regarding the variation within the Eastern European countries, they found that in some countries individual blame is strongly endorsed: especially the citizens of Czech Republic attribute poverty in internal terms (see also Lepianka, 2007). On the other hand, there are countries like Lithuania, Poland, Romania and Croatia in which social blame type of explanation is strongly supported. Finally, social fate type of explanation – living in poverty is explained as simply being part of modern progress – is endorsed especially in Russia.

National case studies from Estonia and Russia have given support to findings of cross-national studies. There is strong support for the idea that the poor have only themselves to blame for their poverty. In addition, both of these countries have strong support for social blame as well. Estonians are as likely to blame the flaws of the economic system as the individualistic reasons. In Russia, on the other hand, the economic system is blamed even more than the individuals themselves (Gorshkov and Tikhonova, 2006; Stephenson, 2000). These results are consistent with Lepianka's (2007: 127–135) comparative analysis of Hungary, Russia, Czech Republic and Estonia as well as with Habibov's (2011) comparative evidence from the Caucasus, Central Asia, Russia and Ukraine. Interesting result is that social blame coincides strongly with individual blame. This illustrates the split-consciousness of public perceptions, which means the coexistence of both dominant and potentially challenging beliefs, where the former represents dominant stratification ideology or value structure of a given country and the latter everyday stratification-related experience of an individual (Kluegel et al., 1995).

Based on qualitative poverty studies in Central Asia, the Caucasus, Ukraine and Moldova, Dudwick et al. (2003: 23) states that, generally, people blamed their own poverty on the failure of the Soviet state and the corruption, indifference, and incompetence of their new leaders. At the same time, however, they often attributed the poverty of others to individual failure, such as laziness, alcoholism, having too many children to support, or having too few children to provide for their old age. Emphazising the positive association between horizontal and vertical trust (see e.g. Rothstein, 2000), they also reported that many of the respondents had become very distrustful of government officials and many suspected that government officials of exploiting the political and economic turmoil for their own gain, at the expense of ordinary citizens. With extensive country selection, the data utilized in this study provides a marvellous possibility to examine the association between attributions for poverty and the context regarding political process, civil liberties and political rights. We can assume that in those countries where the state of democracy is low, citizens blame more likely society for poverty than in those countries where citizens have political freedom and power to influence socio-political issues.

Besides the so called good government, national-level economic conditions have been proved to be associated with welfare attitudes. This is particularly interesting in these times when the Western world is meeting the increasing economic turbulence. In regard to East-West differences as well as variation within post-socialist countries, countries examined in this study have faced global economic crisis in distinct ways. While many Western and Eastern European countries have witnessed economic decline, the economic performance in Western Asia has been strong. Hence, by analyzing the role of changes in economic performance during the period of global financial crisis, the research design of this article provides empirical evidence on the effects of crisis to public perceptions of the causes of poverty. Prior studies have shown that increased financial strain is associated with stronger support for state responsibility for economic provision (Blekesaune, 2007). In regard to explanations of poverty, Gallie and Paugam (2002: 21-24) found that in the majority of European countries, there was a marked decline of support for the individual blame explanation between 1989 and 1993 as economic conditions deteriorated, followed  by an increase between 1993 and 2001 when economic conditions improved. This result is confirmed by Kallio and Niemelä (2013). They found that the larger the economic growth in a country is, more likely people are to blame the poor themselves. Thus, we can assume that changes in country’s economic performance are associated with attributions for poverty regardless of country grouping.

In addition to macro-level economic performance, living conditions in general are found to be linked with attributions for poverty. People living in countries with high unemployment are more likely to endorse social than individualistic explanations (Albrekt Larsen, 2006: 74-79). In addition, the increased incidence of poverty reduces the likelihood of choosing individual laziness as an explanation of poverty (Lepianka, 2007: 118-121). However, there are to some extent contradictory findings as well. While Kallio and Niemelä (2013) did not find association between individual blame type of explanation and incidence of poverty, Lepianka et al. (2010) found out that people living in more developed countries tend to attribute poverty rather to individual and social fate than blame types of explanations. Yet, in contrast to previous studies, the variation between post-socialist countries is far larger than in Western Europe. Consequently, by comparing for instance poorer countries of Central Asia with more well-off Slovenia and Czech Republic, this study provides a possibility to analyze whether the association between attributions for poverty and living conditions is also stronger in the post-socialist countries. This study examines human development index (HDI) as a proxy for living conditions in a given country. Given the great differences in living conditions within the post-socialist space, we assume that in countries of low level of HDI people tend to attribute poverty more with social than individual reasons.

@@

In regard to individual-level determinants, popular attitudes toward the welfare state have traditionally been assumed to be dependent on long-term class-related interest, short-term self-interest and the values and norms that have been internalized by the individuals in the society in question (Taylor-Gooby, 1985; Svallfors, 1995). More recent directions of comparative literature on welfare attitudes tend to also focus on the distribution of marketable capacities such as skill specificity (Iversen and Soskice, 2001) and resource/class-based risks (Cusack et al., 2006; Blomberg et al., 2012; Finseraas and Ringdal, 2012) as key to welfare attitudes. Thus, studies have assumed – and to some extent proved – that groups that have a greater risk of facing social problems or being dependent on welfare state benefits might perceive the government’s role in welfare issues in a distinct way compared to those who are not exposed to these risks.

@@

The previous studies on individual-level determinants of poverty perceptions have focused on individual’s socio-demographic characteristics (e.g. Bullock, 1999; Cozzarelli et al., 2001; Furnham, 1982; Morcöl, 1997; Niemelä, 2008). Results have, however, shown that socio-demographic variables explain very little and they do not reveal clear or consistent patterns. On the other hand, there are studies which support the self-interest hypothesis that those who perceive themselves to be poor are more likely to blame society and social structures for poverty than people well above the poverty line (Niemelä, 2008; Saunders, 2002: 155–156). Yet, there are also contradictory findings indicating that personal experience of poverty might lead to individualistic explanations. For instance, claimants of public welfare seem to hold negative views about other recipients (Bullock, 1999; 2004; Golding and Middleton, 1982: 178).

Studies which have explored the importance of individual’s economic situation on attributions for poverty have focused on individual’s cross-sectional situation of financial circumstances. However, as suggested in the literature on welfare attitudes, not only the risk patterns but also the way people perceive their future risks might have an effect on attitudes toward the welfare state (Andress and Heien, 2001; Boeri et al., 2001; Blomberg et al., 2012; Nygaard Andersen and Ringdal, 2012). In their European comparison Blomberg et al. (2012) found that perceived future risks functioned as an intermediating factor between "objective" risk and welfare state attitudes, i.e. those who perceive their economic future as endangered were more in favour of state responsibility than those who do not expect their future situation to be difficult. Consequently, following the self-interest hypothesis, we can assume that those who perceive their future economic situation to be worse than now endorse more likely social blame type of explanation of poverty than those who do not expect difficulties in their future economic situation.

Not only the current situation or the perceived future but also the past experiences might be associated with individual’s evaluations (e.g. Kumlin, 2007). In regard to attributions for poverty, the effects of past experiences are – according to our knowledge – totally unexplored area of research. In this article we will analyse past experiences by two ways. First, we are interested in whether those whose current economic situation is lower than four years ago perceive the causes of poverty distinct way compared to those whose economic situation has improved or remained in the same level. Second, we will focus on the effects of the global financial crisis. We assume that economic crisis have affected especially those who are the most vulnerable groups in the society. Therefore, we will analyze whether those who report that the economic crisis has affected their household at least a fair amount perceive the causes of poverty differently than those who report that it has had no or only a little effect to their household. According to self-interest hypothesis, we assume that those whose economic situation has declined and those whose household has affected by economic crisis are more inclined to blame society for poverty than those who has not experienced economic problems.

In addition to economic situation, other indicators which measure a greater than average risk of facing more permanent social problems, are "transfer dependency" and low education. Transfer dependency refers to a situation that the respondent’s main source of income relies on private or social transfers (Blomberg et al., 2012; Fridberg and Kangas, 2008). Because of their close relation with welfare system, it can be assumed that welfare recipients perceive more likely social blame than indivual blame type of explanation of poverty compared to the rest of the population.

Low education refers to the lack of capabilities of an individual. Prior findings do not reveal consistent patterns regarding the effects of education. On the one hand, some studies have found the so-called inverted U-relationship which means that people with a middle level of education support the individualistic explanation, whereas people with a low or a high level of education are more likely to endorse structural explanations emphazising social inequalities and the flaws of the economic system of a given country (Feagin, 1972; cf. Furnham, 1982). On the other hand, there are studies which suggest that people with a lower level of education are more likely to favour individualistic explanations. Moreover, when the educational level increases, the likelihood of perceiving causes of poverty in individualistic terms decreases (Niemelä, 2008). It is therefore difficult to form any solid hypotheses as to whether low education as a social risk is associated with perceptions of the causes of poverty in a similar vein than other risk factors discussed above.

@@

Earlier studies have shown that attributions for poverty differ in general between the West and the East of Europe but the variation between East Central European countries is relatively large. There is, however, a shortage of comparative studies which have focused particularly on differences within post-socialist countries. In addition, most of the prior research is based on attitude surveys from the 1990s and most of them have restricted country selection. Therefore, as previous studies have suggested (Kallio and Niemelä, 2013; Lepianka et al., 2010), there is a need for detailed and more up-to-date analysis which would focus on the attributions for poverty within the post-socialist space. Hence, the general purpose of the study is to analyze public perceptions of the causes of poverty in Europe and Central Asia with special emphasis on post-socialist countries.

The study utilizes the data from the second wave of Life in Transition Survey[^***2] (LiTS2), conducted by European Bank of Development and Reconstruction (EBRD) and  World Bank (WB) in 2010. The analyses cover 22 post-socialist countries and 4 Western European countries. The survey was collected using a two-stage clustered stratified sampling procedure to select the households that were included in the sample (for more detailed description of sampling and data collection, see ERBD, 2011). The survey had a special focus on how people’s lives were affected by the global economic crises and its aftermath.

@@

The respondents were asked the following question which measures public perceptions of the causes of poverty: **In your opinion, what is the main reason why there are some people in need in our country today?**

@@

Transfer dependency               Social or private transfers as a main source of income                                                                                  lits

Low education                     No or only compulsory level of education                                                                                                 lits

Perceived low income level        Perceived income level below the country median                                                                                       lits

Income compared to past           Perceived change in income compared to situation in four years ago                                                                    lits

Income compared to future         Expected change in income over the next four years                                                                                    lits

Effect of financial crisis        Whether respondents perceives that their household has suffered the financial crisis great or fair amount                             lits

@@

An additional dataset of selected macro indicators were added to the LiTS2-dataset consisting of three variables: 1) change in total GDP in 2007 - 2010, 2) Human Deveplopment Index in 2010 and 3) Voice and accountability indicator. Data for change in GDP is from World Bank's [World Development Indicators](http://data.worldbank.org/data-catalog/world-development-indicators), Human Development Index from [United Nations Development Programme](http://hdr.undp.org/en/data) and Voice and Accountability from the World Bank's [Worldwide Governance Indicators](http://info.worldbank.org/governance/wgi/index.aspx#home). All macro level indicators are obtained from [Quality of Government Institutes](http://www.qog.pol.gu.se/) data using [rQog](http://markuskainu.fi/rqog/)-package (Kainu, 2014).

The empirical section starts by presenting the results of descriptive analyses. As we assume that individual level perceptions are dependent both on individual and country-level contextual factors and we have a hierarchical two level dataset, it is recommended to use multilevel regression analysis for this task (Snijders and Bosker, 2011) as multilevel regression analysis allows the estimation of country-level effects on individual level outcomes. The analysis is performed using glmer-function from lme4 package in R. All of the models employed are so called multilevel logistic random intercept models at the level of individuals and countries. Model summary tables include both a fixed effects and a random effects part. The fixed effects part will include odds ratios and statistical significances of independent variables, while the random effects part will indicate the country-level variance and standard deviation, deviance and number of cases and groups.

We are modelling the date in stepwise manner first running the empty models for both social blame (Table 3) and individual blame (Table 4) type of explanation. At second stage we include all individual-level predictors and in third stage we include only the macro-level predictors. In final models we include both the individual and macro-level variables. As we are using *logistics regression* instead of *multinomial logistic regression* with dependent variable of nominal scale, we confirm our findings by running a further model where we predict the support for social blame against the support for individual blame using both the individual and contextual level variables. This model is compared with full models from Tables 3 and 4 in Appendix Table 1.

@@

In regard to differences between the East and the West of Europe, results show that there are some differences. Yet, there are also substantial variation within Western Europe, which correspond earlier findings on cross-national differences in Western Europe (Lepianka et al., 2010; Kallio and Niemelä, 2010). Clearest differences between the West and the East is the stronger support for social fate in Western Europe. In Germany and Sweden external reasons for poverty receive the majority of support. On the other hand, attributions for poverty in France and Italy seems to be quite similar with post-socialist countries - blame types of explanations are endorsed more likely than fate types of explanations. Finally, the figures of Great Britain are the example of selective welfare policy dominated liberal regime in which the individual blame type of explanations is strongly supported (see also Albrekt Larsen, 2006).

>>>>

As Joseph Schumpeter (1942, p. 12) once argued, "attitudes are coins that do not readily melt". This means that normative expectations, justice beliefs or welfare perceptions are very hard to change. In this way, welfare attitudes can be associated with policy changes by highlighting public sentiments, which are normative background assumptions or ideas that constrain decision-making and institutional change by limiting the range of programme that decision-making elites are likely to perceive as acceptable and legitimate to both their constituents and themselves (Campbell 2004, pp. 96-100). On the other hand, "the worlds on welfare" literature on welfare attitudes has in turn emphasized that type of welfare policy affects public attitudes (Andress and Heien 2001; Arts and Gelissen 2001; Jaeger 2006; Svallfors 1997). In regard to public perceptions of the causes of poverty, in the marginal welfare-policy dominated liberal regime, the poor will be asked to fulfil much harder deservingness criteria than in a universal social democratic regime (van Oorschot 2006). In countries, in which welfare programmes are more marginal and less generous, explanations of poverty are more often based on individual morality, reflecting the strength of the dominant individualistic ideology, than they are in the Continental and Nordic welfare states – where external and structural explanations prevail (Albrekt Larsen 2006, pp. 47-56; Alesina and Gleaser 2004).

The focus of this study is on the public perceptions of the causes of poverty in post-socialist countries. Since the early 1990s, the development of post-socialist welfare state has received abundant scholarly attention (e.g. Cerami and Vanhuysse 2009; Inglot 2008; Haggard and Kauffman 2008; Cook 2007; Aidukaite 2004; Deacon 1992). As the decades under socialist regime have had an undeniable influence on societies, these studies have emphasized the role of institutional differences within region as well as pre-socialist historical origins of the welfare states. Now after 20 years of transition, as there has not been a visible convergence towards a common "post-socialist model", the attention has shifted even more on various institutional and historical explanations (Szikra and Tomka 2009; Tomka 2006; Cerami and Vanhuysse 2009; Inglot 2008).

The purpose of this article is to examine cross-national differences in attributions for poverty in Europe and Western Asia with special emphasis on post-socialist countries. Prior findings of the comparative studies on public perceptions of the causes of poverty have emphasized clear differences between Western and Eastern European countries (Gallie and Paugam 2002; Kallio and Niemelä 2014; Lepianka 2007; van Oorschot and Halman 2000). East Europeans are more inclined to blame either the individual behavior or the flaws of the economic system of country than their Western counterparts. Hence, it seems that unlike in Western Europe traditionally opposite attitudinal dimensions, i.e. structural and individualistic explanations of poverty coincide with each other in East Europeans' perceptions. However, there is shortage of comparative studies on attributions for poverty in particular focusing on the East Central European countries or on the whole post-socialist space. There is therefore a need for a more detailed and up-to-date analysis of the public perceptions of the causes of poverty in post-socialist countries.

The novelty of this study is that it covers wide range of countries from Europe and Western Asia. Thus, compared to previous studies, it provides more valid evidence on the differences in perceptions between the East and the West and within post-socialist regime as well as on the importance of the post-socialist legacy. In addition, the study applies multilevel techniques in order to analyze the role of individual and contextual level explanatory factors as determinants of poverty attributions. The article explores to what extent specific risks related to household financial circumstances at the individual-level and the state of democracy, income inequality and changes in national economy at the contextual level are associated with different types of lay explanations of poverty. Finally, extensive country-selection of our data provides a novel opportunity to examine the effects of the financial crisis both at the individual and contextual level and their association with attributions for poverty. At the individual-level, the study focuses not only on the individuals’ current financial situation. Instead, the article provides an empirical example of how past experiences, current financial position as well as future expectations concerning individual’s economic situation are associated with perceptions of the causes of poverty.

@@

Pioneering empirical studies on lay explanations of poverty emerged at the early 1970s when Joe R. Feagin (1972; 1975) examined Americans' perceptions of the causes of poverty. He categorized reasons for poverty into three basic categories: individualistic, structural and fatalistic. Later, many factor analytic studies have given empirical support to Feagin’s categorization (e.g. Feather 1974; Furnham 1982; Hunt 1996; Morçöl 1997; Nasser et al. 2002; Niemelä 2008). Yet some studies have expanded our understanding by incorporating more contemporary beliefs into the attributional scales. These works emphasize especially the relevance of cultural attributions for poverty (Bullock et al. 2003; Cozzarelli et al. 2001).

@@

Prior empirical findings on attributions for poverty in Europe have shown that there are some differences between the West and the East of Europe. Western Europeans are more inclined to support social explanations for poverty than their Eastern counterparts. On the other hand, individual blame type of explanation is generally more popular in the East than in the West of Europe (Kallio and Niemelä 2014; Kreidl 2000; Lepianka 2007; Lepianka et al. 2010; van Oorschot and Halman 2000). Earlier results suggest, however, that there are variations both within and across groups of countries. Interestingly enough, the variation is highest within the East Central European countries (Kallio and Niemelä 2014). Also Lepianka et al. (2010) come to the conclusion that further investigation is required due to rather large variation across countries. They showed that individual fate type of explanation is supported relatively less frequently in the East of Europe than in Western European countries. Regarding the variation within the Eastern European countries, they found that in some countries individual blame is strongly endorsed: especially the citizens of Czech Republic attribute poverty in internal terms (see also Lepianka 2007). On the other hand, there are countries like Lithuania, Poland, Romania and Croatia in which social blame type of explanation is strongly supported. Finally, social fate type of explanation – living in poverty is explained as simply being part of modern progress – is endorsed especially in Russia.

National case studies from Estonia and Russia have given support to findings of cross-national studies. There is strong support for the idea that the poor have only themselves to blame for their poverty. In addition, both of these countries have strong support for social blame as well. Estonians are as likely to blame the flaws of the economic system as the individualistic reasons. In Russia, on the other hand, the economic system is blamed even more than the individuals themselves (Gorshkov and Tikhonova 2006; Stephenson 2000). These results are consistent with Lepianka's (2007, pp. 127–135) comparative analysis of Hungary, Russia, Czech Republic and Estonia as well as with Habibov's (2011) comparative evidence from the Caucasus, Central Asia, Russia and Ukraine. Interesting result is that social blame coincides strongly with individual blame. This illustrates the split-consciousness of public perceptions, which means the coexistence of both dominant and potentially challenging beliefs, where the former represents dominant stratification ideology or value structure of a given country and the latter everyday stratification-related experience of an individual (Kluegel et al. 1995).

Based on qualitative poverty studies in Central Asia, the Caucasus, Ukraine and Moldova, Dudwick et al. (2003, p. 23) states that, generally, people blamed their own poverty on the failure of the Soviet state and the corruption, indifference, and incompetence of their new leaders. At the same time, however, they often attributed the poverty of others to individual failure, such as laziness, alcoholism, having too many children to support, or having too few children to provide for their old age. Emphazising the positive association between horizontal and vertical trust (see e.g. Rothstein 2000), they also reported that many of the respondents had become very distrustful of government officials and many suspected that government officials of exploiting the political and economic turmoil for their own gain, at the expense of ordinary citizens. With extensive country selection, the data utilized in this study provides a marvellous possibility to examine the association between attributions for poverty and the context regarding political process, civil liberties and political rights. We can assume that in those countries where the state of democracy is low, citizens blame more likely society for poverty than in those countries where citizens have political freedom and power to influence socio-political issues.

Besides the so called good government, national-level economic conditions have been proved to be associated with welfare attitudes. This is particularly interesting in these times when the Western world is meeting the increasing economic turbulence. In regard to East-West differences as well as variation within post-socialist countries, countries examined in this study have faced global economic crisis in distinct ways. While many Western and Eastern European countries have witnessed economic decline, the economic performance in Western Asia has been strong. Hence, by analyzing the role of changes in economic performance during the period of global financial crisis, the research design of this article provides empirical evidence on the effects of crisis to public perceptions of the causes of poverty. Prior studies have shown that increased financial strain is associated with stronger support for state responsibility for economic provision (Blekesaune 2007). In regard to explanations of poverty, Gallie and Paugam (2002, pp. 21-24) found that in the majority of European countries, there was a marked decline of support for the individual blame explanation between 1989 and 1993 as economic conditions deteriorated, followed  by an increase between 1993 and 2001 when economic conditions improved. This result is confirmed by Kallio and Niemelä (2014). They found that the larger the economic growth in a country is, more likely people are to blame the poor themselves. Thus, we can assume that changes in country’s economic performance are associated with attributions for poverty regardless of country grouping.

In addition to macro-level economic performance, living conditions in general are found to be linked with attributions for poverty. People living in countries with high unemployment are more likely to endorse social than individualistic explanations (Albrekt Larsen 2006, pp. 74-79). In addition, the increased incidence of poverty reduces the likelihood of choosing individual laziness as an explanation of poverty (Lepianka, 2007: 118-121). However, there are to some extent contradictory findings as well. While Kallio and Niemelä (2014) did not find association between individual blame type of explanation and incidence of poverty, Lepianka et al. (2010) found out that people living in more developed countries tend to attribute poverty rather to individual and social fate than blame types of explanations. Yet, in contrast to previous studies, the variation between post-socialist countries is far larger than in Western Europe. Consequently, by comparing for instance poorer countries of Central Asia with more well-off Slovenia and Czech Republic, this study provides a possibility to analyze whether the association between attributions for poverty and living conditions is also stronger in the post-socialist countries. This study examines Human Development Index (HDI) as a proxy for living conditions in a given country. Given the great differences in living conditions within the post-socialist space, we assume that in countries of low level of HDI people tend to attribute poverty more with social than individual reasons.

@@

In regard to individual-level determinants, popular attitudes toward the welfare state have traditionally been assumed to be dependent on long-term class-related interest, short-term self-interest and the values and norms that have been internalized by the individuals in the society in question (Taylor-Gooby 1985; Svallfors 2012). More recent directions of comparative literature on welfare attitudes tend to also focus on the distribution of marketable capacities such as skill specificity (Iversen and Soskice 2001) and resource/class-based risks (Cusack et al. 2006; Blomberg et al. 2012; Finseraas and Ringdal 2012) as key to welfare attitudes. Thus, studies have assumed – and to some extent proved – that groups that have a greater risk of facing social problems or being dependent on welfare state benefits might perceive the government’s role in welfare issues in a distinct way compared to those who are not exposed to these risks.

@@

The previous studies on individual-level determinants of poverty perceptions have focused on individual’s socio-demographic characteristics (e.g. Bullock 1999; Cozzarelli et al. 2001; Furnham 1982; Morçöl 1997; Niemelä 2008). Results have, however, shown that socio-demographic variables explain very little and they do not reveal clear or consistent patterns. On the other hand, there are studies which support the self-interest hypothesis that those who perceive themselves to be poor are more likely to blame society and social structures for poverty than people well above the poverty line (Niemelä 2008; Saunders 2003). Yet, there are also contradictory findings indicating that personal experience of poverty might lead to individualistic explanations. For instance, claimants of public welfare seem to hold negative views about other recipients (Bullock 1999; 2004; Golding and Middleton 1982, p. 178).

Studies which have explored the importance of individual’s economic situation on attributions for poverty have focused on individual’s cross-sectional situation of financial circumstances. However, as suggested in the literature on welfare attitudes, not only the risk patterns but also the way people perceive their future risks might have an effect on attitudes toward the welfare state (Andress and Heien 2001; Boeri et al. 2001; Blomberg et al. 2012; Nygaard Andersen and Ringdal 2012). In their European comparison Blomberg et al. (2012) found that perceived future risks functioned as an intermediating factor between "objective" risk and welfare state attitudes, i.e. those who perceive their economic future as endangered were more in favour of state responsibility than those who do not expect their future situation to be difficult. Consequently, following the self-interest hypothesis, we can assume that those who perceive their future economic situation to be worse than now endorse more likely social blame type of explanation of poverty than those who do not expect difficulties in their future economic situation.

Not only the current situation or the perceived future but also the past experiences might be associated with individual’s evaluations (e.g. Kumlin 2007). In regard to attributions for poverty, the effects of past experiences are – according to our knowledge – totally unexplored area of research. In this article we will analyze past experiences by two ways. First, we are interested in whether those whose current economic situation is lower than four years ago perceive the causes of poverty distinct way compared to those whose economic situation has improved or remained in the same level. Second, we will focus on the effects of the global financial crisis. We assume that economic crisis have affected especially those who are the most vulnerable groups in the society. Therefore, we will analyze whether those who report that the economic crisis has affected their household at least a fair amount perceive the causes of poverty differently than those who report that it has had no or only a little effect to their household. According to self-interest hypothesis, we assume that those whose economic situation has declined and those whose household has affected by economic crisis are more inclined to blame society for poverty than those who has not experienced economic problems.

In addition to economic situation, other indicators which measure a greater than average risk of facing more permanent social problems, are "transfer dependency" and low education. Transfer dependency refers to a situation that the respondent’s main source of income relies on private or social transfers (Blomberg et al. 2012; Fridberg and Kangas 2008). Because of their close relation with welfare system, it can be assumed that welfare recipients perceive more likely social blame than indivual blame type of explanation of poverty compared to the rest of the population.

Low education refers to the lack of capabilities of an individual. Prior findings do not reveal consistent patterns regarding the effects of education. On the one hand, some studies have found the so-called inverted U-relationship which means that people with a middle level of education support the individualistic explanation, whereas people with a low or a high level of education are more likely to endorse structural explanations emphazising social inequalities and the flaws of the economic system of a given country (Feagin 1972; cf. Furnham 1982). On the other hand, there are studies which suggest that people with a lower level of education are more likely to favour individualistic explanations. Moreover, when the educational level increases, the likelihood of perceiving causes of poverty in individualistic terms decreases (Niemelä 2008). It is therefore difficult to form any solid hypotheses as to whether low education as a social risk is associated with perceptions of the causes of poverty in a similar vein than other risk factors discussed above.

@@

Earlier studies have shown that attributions for poverty differ in general between the West and the East of Europe but the variation between East Central European countries is relatively large. There is, however, a shortage of comparative studies which have focused particularly on differences within post-socialist countries. In addition, most of the prior research is based on attitude surveys from the 1990s and most of them have restricted country selection. Therefore, as previous studies (Kallio and Niemelä 2014; Lepianka et al. 2010) have suggested, there is a need for detailed and more up-to-date analysis which would focus on the attributions for poverty within the post-socialist space. Hence, the general purpose of the study is to analyze public perceptions of the causes of poverty in Europe and Central Asia with special emphasis on post-socialist countries.

The study utilizes the data from the second wave of Life in Transition Survey[^***2] (LiTS2), conducted by European Bank of Development and Reconstruction (EBRD) and  World Bank in 2010. The analyses cover 22 post-socialist countries and 4 Western European countries. The survey was collected using a two-stage clustered stratified sampling procedure to select the households that were included in the sample (for more detailed description of sampling and data collection, see ERBD 2011). The survey had a special focus on how people’s lives were affected by the global economic crises and its aftermath.

@@

The respondents were asked the following question which measures public perceptions of the causes of poverty: In your opinion, what is the main reason why there are some people in need in our country today?

@@

Transfer dependency               Social or private transfers as a main source of income                                                                                  LiTS2

Low education                     No or only compulsory level of education                                                                                                 LiTS2

Perceived low income level        Perceived income level below the country median                                                                                       LiTS2

Income compared to past           Perceived change in income compared to situation in four years ago                                                                    LiTS2

Income compared to future         Expected change in income over the next four years                                                                                    LiTS2

Effect of financial crisis        Whether respondents perceives that their household has suffered the financial crisis great or fair amount                             LiTS2

@@

An additional dataset of selected macro indicators were added to the LiTS2-dataset consisting of three variables: 1) change in total GDP in 2007 - 2010, 2) Human Deveplopment Index in 2010 and 3) Voice and accountability indicator. Data for change in GDP is from World Bank's [World Development Indicators](http://data.worldbank.org/data-catalog/world-development-indicators), Human Development Index from [United Nations Development Programme](http://hdr.undp.org/en/data) and Voice and Accountability from the World Bank's [Worldwide Governance Indicators](http://info.worldbank.org/governance/wgi/index.aspx#home). All macro level indicators are obtained from [Quality of Government Institutes](http://www.qog.pol.gu.se/) data using [rQog](http://markuskainu.fi/rqog/)-package (Kainu 2014).

The empirical section starts by presenting the results of descriptive analyses. As we assume that individual level perceptions are dependent both on individual and country-level contextual factors and we have a hierarchical two level dataset, it is recommended to use multilevel regression analysis for this task (Snijders and Bosker 2011) as multilevel regression analysis allows the estimation of country-level effects on individual level outcomes. The analysis is performed using glmer-function from lme4 package in R. All of the models employed are so called multilevel logistic random intercept models at the level of individuals and countries. Model summary tables include both a fixed effects and a random effects part. The fixed effects part will include odds ratios and statistical significances of independent variables, while the random effects part will indicate the country-level variance and standard deviation, deviance and number of cases and groups.

We are modelling the date in stepwise manner first running the empty models for both social blame (Table 3) and individual blame (Table 4) type of explanation. At the second stage we include all individual-level predictors and in the third stage we include only the macro-level predictors. In final models we include both the individual and macro-level variables. As we are using *logistics regression* instead of *multinomial logistic regression* with dependent variable of nominal scale, we confirm our findings by running a further model where we predict the support for social blame against the support for individual blame using both the individual and contextual level variables. This model is compared with full models from Tables 3 and 4 in Appendix Table 1.

@@

In regard to differences between the East and the West of Europe, results show that there are some differences. Yet, there are also substantial variation within Western Europe, which correspond earlier findings on cross-national differences in Western Europe (Lepianka et al. 2010; Kallio and Niemelä 2014). Clearest differences between the West and the East is the stronger support for social fate in Western Europe. In Germany and Sweden external reasons for poverty receive the majority of support. On the other hand, attributions for poverty in France and Italy seems to be quite similar with post-socialist countries - blame types of explanations are endorsed more likely than fate types of explanations. Finally, the figures of Great Britain are the example of selective welfare policy dominated liberal regime in which the individual blame type of explanations is strongly supported (see also Albrekt Larsen 2006).
